A pipeline to analyse time-course gene expression data

### 2020-07-27 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

\begin{abstract}
The phenotypic diversity of cells is governed by a complex equilibrium between their genetic identity and their environmental interactions: Understanding the dynamics of gene expression is a fundamental question of biology. However, analysing time-course transcriptomic data raises unique challenging statistical and computational questions, requiring the development of novel methods and software. Using as case study time-course transcriptomics data from mice exposed to different strains of influenza, this workflow provides a step-by-step tutorial of the methodology used to analyse time-course data: (1) quality control and normalization of the data set; (2) differential expression analysis using functional data analysis; (3) clustering of time-course data; (4) interpreting clusters with GO term and KEGG pathway enrichment analysis.
\end{abstract}

@@

The manuscript was last compile on 2020-07-22 10:25:43.

@@

While there are many methods that have been proposed for discrete aspects of
time course data, the entire workflow for analysis of such data remains
difficult, particularly for long, developmental time series. Most methods
proposed for time course data are concerned with detecting genes that are
changing over time (differential expression analysis), examples being \texttt{edge}
\citep{storey:significance}, functional component analysis based models \citep{wu:more},
time-course permutation tests \citep{park:statistical}, and multiple testing
strategies to combine single time point differential expression analysis
\citep{sun:multiple}. However, with long time course data sets, particularly in
developmental systems, a massive number of genes will show \emph{some} change. For
example, in a study of mice lung tissues infected with influenza that we
consider in this workflow, over 50\% of genes
are shown to be changing over time. The task in these settings is often not to
detect changes in genes, but to categorize them into biologically interpretable
patterns.

@@

While we showcase this pipeline on micro-array data, \citep{varoquaux:lifecycle}
leverages a similar set of steps to analyse RNA-seq data of the lifetime
transcriptomic response of the crop \emph{S. bicolor} to drought.

@@

\begin{center}\includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/topPvalueGenes-1} \end{center}

@@

\begin{center}\includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/topLogFold-1} \end{center}

@@

\begin{center}\includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/unnamed-chunk-18-1} \end{center}

@@

\begin{center}\includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/genesVsCentroids-1} \end{center}

@@

We can visualize the impact of this filtering process by considering the distribution of goodness-of-fit scores for each
cluster if we did not have the filtering cutoff of \texttt{splines\_kmeans\_score\_and\_label}, i.e.~if we simply assign every gene to a cluster based on which gives them their minimum score.
We display for each cluster, the scores of the genes assigned to that cluster (Figure \ref{visualizeScores}), as compared to the cutoff value for what that score must be under our filtering procedure. We can do this by rerunning \texttt{splines\_kmeans\_score\_and\_label} and setting the filter \texttt{percentage\_genes\_to\_label=1} (and we can speed up the calculation by giving our previously calculated score via the argument \texttt{previous\_scores})

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# order them by their score in cluster}
\NormalTok{ord =}\StringTok{ }\KeywordTok{order}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores[genes_in_cluster2,}
\NormalTok{                     cluster_of_interest])}
\NormalTok{cluster2Score =}\StringTok{ }\NormalTok{genes_in_cluster2[ord]}
\NormalTok{dataToPlot =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(}\KeywordTok{assay}\NormalTok{(moanin_model[cluster2Score[}\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{], ]),}
    \StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[cluster_of_interest, ])}
\KeywordTok{plot_splines_data}\NormalTok{(moanin_model, }\DataTypeTok{data=}\NormalTok{dataToPlot, }
     \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}\DataTypeTok{simpleY =}\OtherTok{FALSE}\NormalTok{,}
     \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

First, we select the genes of interest, namely those in cluster 8, based on
our goodness-of-fit assignment.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cluster_to_plot =}\StringTok{ }\DecValTok{8}
\NormalTok{genes_to_plot =}\StringTok{  }\KeywordTok{names}\NormalTok{(labels[labels }\OperatorTok{==}\StringTok{ }\NormalTok{cluster_to_plot])}
\end{Highlighting}
\end{Shaded}

Now we will create the heatmaps of these genes (1283 genes, Figure \ref{fig:heatmapCluster8}).

@@

We plot the WCSS for \(k=2-20\) here (Figure \ref{fig:elbow_method}).
We see that as expected the WCSS continues
to drop, but there is no clear drop in the decrease, except for very small
values of \(k\) (3-4 clusters). However, 3-4 seems a very small number of gene
clusters to find, given the complexity

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{consensus_matrix_stability_}\DecValTok{5}\NormalTok{ =}\StringTok{ }\KeywordTok{consensus_matrix}\NormalTok{(stability_}\DecValTok{5}\NormalTok{,}
                        \DataTypeTok{scale=}\OtherTok{FALSE}\NormalTok{)}
\KeywordTok{aheatmap}\NormalTok{(consensus_matrix_stability_}\DecValTok{5}\NormalTok{[}\DecValTok{1}\OperatorTok{:}\DecValTok{1000}\NormalTok{, }\DecValTok{1}\OperatorTok{:}\DecValTok{1000}\NormalTok{], }\DataTypeTok{Rowv=}\OtherTok{FALSE}\NormalTok{,}
     \DataTypeTok{Colv=}\OtherTok{FALSE}\NormalTok{,}
     \DataTypeTok{treeheight=}\DecValTok{0}\NormalTok{, }\DataTypeTok{main=}\StringTok{"K=5"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\caption{Consensus matrix of proportion of times samples were in the same cluster over bootstrap samples for k=5 and 20.}\label{fig:bootstrapConsensus-1}
\end{figure}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{consensus_matrix_stability_}\DecValTok{20}\NormalTok{ =}\StringTok{ }\KeywordTok{consensus_matrix}\NormalTok{(stability_}\DecValTok{20}\NormalTok{,}
                         \DataTypeTok{scale=}\OtherTok{FALSE}\NormalTok{)}
\KeywordTok{aheatmap}\NormalTok{(consensus_matrix_stability_}\DecValTok{20}\NormalTok{[}\DecValTok{1}\OperatorTok{:}\DecValTok{1000}\NormalTok{, }\DecValTok{1}\OperatorTok{:}\DecValTok{1000}\NormalTok{], }\DataTypeTok{Rowv=}\OtherTok{FALSE}\NormalTok{,}
     \DataTypeTok{Colv=}\OtherTok{FALSE}\NormalTok{,}
     \DataTypeTok{treeheight=}\DecValTok{0}\NormalTok{, }\DataTypeTok{main=}\StringTok{"K=20"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/bootstrapConsensus-2}

\caption{Consensus matrix of proportion of times samples were in the same cluster over bootstrap samples for k=5 and 20.}\label{fig:bootstrapConsensus-2}
\end{figure}

@@

\CommentTok{## Get gene names of genes in Cluster 8}
\NormalTok{cluster =}\StringTok{ }\DecValTok{8}
\NormalTok{labels =}\StringTok{ }\KeywordTok{unlist}\NormalTok{(labels)}
\NormalTok{gene_names =}\StringTok{ }\KeywordTok{names}\NormalTok{(labels)}
\NormalTok{genes =}\StringTok{ }\NormalTok{gene_names[labels }\OperatorTok{==}\StringTok{ }\NormalTok{cluster]}

@@

Then we use the function \texttt{find\_enriched\_pathway} in the \texttt{KEGGprofile} package to determine whether any KEGG pathways are enriched in cluster 8, i.e.~whether a higher percentage of genes from a single pathway are found in cluster 8 than we would expect by simply proportional assignment of genes to clusters.

@@

\begin{tabular}{llrr}
\toprule
  & Pathway & Percentage & Adj. p-value\\
\midrule
04740 & Olfactory transduction & 8 & 0\\
04080 & Neuroactive ligand-receptor interaction & 10 & 0\\
\bottomrule
\end{tabular}

\textbf{EAP: I changed this to be the min(10, nrow(pathways)) but wondering if there's a problem, since presumably there used to be more than 10 pathways found significant and on my machine there are only 2?}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Create logical vector of whether in cluster 8}
\NormalTok{assignments =}\StringTok{ }\NormalTok{labels }\OperatorTok{==}\StringTok{ }\NormalTok{cluster}

@@

\begin{tabular}{llrrrlr}
\toprule
GO ID & Description & Annotated & Significant & Expected & P-value & Adj. p-value\\
\midrule
GO:0044260 & cellular macromolecule metabolic process & 3762 & 3607 & 3540.21 & 0.00031 & 0.276\\
GO:0043603 & cellular amide metabolic process & 467 & 458 & 439.47 & 0.00036 & 0.276\\
GO:0043161 & proteasome-mediated ubiquitin-dependent ... & 201 & 199 & 189.15 & 0.00042 & 0.276\\
GO:0016071 & mRNA metabolic process & 285 & 280 & 268.20 & 0.00052 & 0.276\\
GO:0070925 & organelle assembly & 427 & 417 & 401.83 & 0.00052 & 0.276\\
\addlinespace
GO:0019058 & viral life cycle & 123 & 123 & 115.75 & 0.00054 & 0.276\\
GO:0018130 & heterocycle biosynthetic process & 1809 & 1733 & 1702.35 & 0.00093 & 0.348\\
GO:0019438 & aromatic compound biosynthetic process & 1810 & 1734 & 1703.29 & 0.00098 & 0.348\\
GO:0042110 & T cell activation & 273 & 268 & 256.91 & 0.00102 & 0.348\\
GO:0006954 & inflammatory response & 410 & 398 & 385.83 & 0.00104 & 0.348\\
\addlinespace
GO:0006163 & purine nucleotide metabolic process & 199 & 198 & 187.27 & 0.00108 & 0.348\\
GO:0051301 & cell division & 300 & 294 & 282.31 & 0.00109 & 0.348\\
GO:0070507 & regulation of microtubule cytoskeleton o... & 110 & 110 & 103.51 & 0.00120 & 0.361\\
GO:0006091 & generation of precursor metabolites and ... & 222 & 220 & 208.91 & 0.00138 & 0.392\\
GO:0006364 & rRNA processing & 106 & 106 & 99.75 & 0.00153 & 0.412\\
\bottomrule
\end{tabular}

@@

##  [3] ggfortify_0.4.10            timecoursedata_0.99.13     

>>>>

\begin{abstract}
The phenotypic diversity of cells is governed by a complex equilibrium between their genetic identity and their environmental interactions: Understanding the dynamics of gene expression is a fundamental question of biology. However, analysing time-course transcriptomic data raises unique challenging statistical and computational questions, requiring the development of novel methods and software. Using as case study time-course transcriptomics data from mice exposed to different strains of influenza, this workflow provides a step-by-step tutorial of the methodology used to analyse time-course data: (1) quality control and normalization of the dataset; (2) differential expression analysis using functional data analysis; (3) clustering of time-course data; (4) interpreting clusters with GO term and KEGG pathway enrichment analysis.
\end{abstract}

@@

The manuscript was last compile on 2020-07-27 16:08:41.

@@

While there are many methods that have been proposed for discrete aspects of
time course data, the entire workflow for analysis of such data remains
difficult, particularly for long, developmental time series. Most methods
proposed for time course data are concerned with detecting genes that are
changing over time (differential expression analysis), examples being \texttt{edge}
\citep{storey:significance}, functional component analysis based models \citep{wu:more},
time-course permutation tests \citep{park:statistical}, and multiple testing
strategies to combine single time point differential expression analysis
\citep{sun:multiple}. However, with long time course datasets, particularly in
developmental systems, a massive number of genes will show \emph{some} change. For
example, in a study of mice lung tissues infected with influenza that we
consider in this workflow, over 50\% of genes
are shown to be changing over time. The task in these settings is often not to
detect changes in genes, but to categorize them into biologically interpretable
patterns.

@@

While we showcase this pipeline on micro-array data,
\citep{varoquaux:transcriptomic}
leverages a similar set of steps to analyse RNA-seq data of the lifetime
transcriptomic response of the crop \emph{S. bicolor} to drought.

@@

{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/topPvalueGenes-1}

\caption{Top 10 genes with the smallest p-values}\label{fig:topPvalueGenes}
\end{figure}

@@

{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/topLogFold-1}

\caption{Top 10 genes with largest absolute timecourse log-fold change}\label{fig:topLogFold}
\end{figure}

@@

{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/unnamed-chunk-18-1}

\caption{K-means centroids}\label{fig:unnamed-chunk-18}
\end{figure}

@@

{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/genesVsCentroids-1}

\caption{Genes from cluster 2}\label{fig:genesVsCentroids}
\end{figure}

@@

We can visualize the impact of this filtering process by considering the
distribution of goodness-of-fit scores for each cluster if we did not have the
filtering cutoff of \texttt{splines\_kmeans\_score\_and\_label}, i.e.~if we simply assign
every gene to a cluster based on which gives them their minimum score. We
display for each cluster, the scores of the genes assigned to that cluster
(Figure \ref{fig:visualizeScores}), as compared to the cutoff value for what
that score must be under our filtering procedure. We can do this by rerunning
\texttt{splines\_kmeans\_score\_and\_label} and setting the filter
\texttt{percentage\_genes\_to\_label=1} (and we can speed up the calculation by giving
our previously calculated score via the argument \texttt{previous\_scores})

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# order them by their score in cluster}
\NormalTok{ord =}\StringTok{ }\KeywordTok{order}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores[genes_in_cluster2,}
\NormalTok{                     cluster_of_interest])}
\NormalTok{cluster2Score =}\StringTok{ }\NormalTok{genes_in_cluster2[ord]}
\NormalTok{dataToPlot =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(}
    \KeywordTok{assay}\NormalTok{(moanin_model[cluster2Score[}\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{], ]),}
    \StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[cluster_of_interest, ])}
\KeywordTok{plot_splines_data}\NormalTok{(}
\NormalTok{    moanin_model, }\DataTypeTok{data=}\NormalTok{dataToPlot, }
    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{, }\DataTypeTok{simpleY=}\OtherTok{FALSE}\NormalTok{,}
    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{, }\FloatTok{2.5}\NormalTok{, }\DecValTok{2}\NormalTok{, }\FloatTok{0.1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

First, we select the genes of interest, namely those in cluster 6, based on
our goodness-of-fit assignment.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cluster_to_plot =}\StringTok{ }\DecValTok{6}
\NormalTok{genes_to_plot =}\StringTok{  }\KeywordTok{names}\NormalTok{(labels[labels }\OperatorTok{==}\StringTok{ }\NormalTok{cluster_to_plot])}
\end{Highlighting}
\end{Shaded}

Now we will create the heatmaps of these genes (3746 genes, Figure \ref{fig:heatmapCluster8}).

@@

We plot the WCSS for \(k=2-20\) here (Figure \ref{fig:elbowMethod}).
We see that as expected the WCSS continues
to drop, but there is no clear drop in the decrease, except for very small
values of \(k\) (3-4 clusters). However, 3-4 seems a very small number of gene
clusters to find, given the complexity

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow=}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{))}
\NormalTok{consensus_matrix_stability_}\DecValTok{5}\NormalTok{ =}\StringTok{ }\KeywordTok{consensus_matrix}\NormalTok{(stability_}\DecValTok{5}\NormalTok{,}
                        \DataTypeTok{scale=}\OtherTok{FALSE}\NormalTok{)}
\KeywordTok{aheatmap}\NormalTok{(consensus_matrix_stability_}\DecValTok{5}\NormalTok{[}\DecValTok{1}\OperatorTok{:}\DecValTok{1000}\NormalTok{, }\DecValTok{1}\OperatorTok{:}\DecValTok{1000}\NormalTok{], }\DataTypeTok{Rowv=}\OtherTok{FALSE}\NormalTok{,}
     \DataTypeTok{Colv=}\OtherTok{FALSE}\NormalTok{,}
     \DataTypeTok{treeheight=}\DecValTok{0}\NormalTok{, }\DataTypeTok{main=}\StringTok{"K=5"}\NormalTok{)}

\NormalTok{consensus_matrix_stability_}\DecValTok{20}\NormalTok{ =}\StringTok{ }\KeywordTok{consensus_matrix}\NormalTok{(stability_}\DecValTok{20}\NormalTok{,}
                         \DataTypeTok{scale=}\OtherTok{FALSE}\NormalTok{)}
\KeywordTok{aheatmap}\NormalTok{(consensus_matrix_stability_}\DecValTok{20}\NormalTok{[}\DecValTok{1}\OperatorTok{:}\DecValTok{1000}\NormalTok{, }\DecValTok{1}\OperatorTok{:}\DecValTok{1000}\NormalTok{], }\DataTypeTok{Rowv=}\OtherTok{FALSE}\NormalTok{,}
     \DataTypeTok{Colv=}\OtherTok{FALSE}\NormalTok{,}
     \DataTypeTok{treeheight=}\DecValTok{0}\NormalTok{, }\DataTypeTok{main=}\StringTok{"K=20"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\caption{Consensus matrix of proportion of times samples were in the same cluster over bootstrap samples for k=5 and 20.}\label{fig:bootstrapConsensus}
\end{figure}

@@

\CommentTok{## Get gene names of genes in Cluster 8}
\NormalTok{cluster =}\StringTok{ }\DecValTok{6}
\NormalTok{gene_names =}\StringTok{ }\KeywordTok{names}\NormalTok{(labels)}
\NormalTok{genes =}\StringTok{ }\NormalTok{gene_names[labels }\OperatorTok{==}\StringTok{ }\NormalTok{cluster]}

@@

Then we use the function \texttt{find\_enriched\_pathway} in the \texttt{KEGGprofile} package
to determine whether any KEGG pathways are enriched in cluster 6, i.e.~whether
a higher percentage of genes from a single pathway are found in cluster 6 than
we would expect by simply proportional assignment of genes to clusters.

@@

\begin{tabular}{llrr}
\toprule
  & Pathway & Percentage & Adj. p-value\\
\midrule
04060 & Cytokine-cytokine receptor interaction & 31 & 0\\
04620 & Toll-like receptor signaling pathway & 39 & 0\\
04621 & NOD-like receptor signaling pathway & 45 & 0\\
04623 & Cytosolic DNA-sensing pathway & 45 & 0\\
04630 & Jak-STAT signaling pathway & 28 & 0\\
\addlinespace
04210 & Apoptosis & 35 & 0\\
04380 & Osteoclast differentiation & 31 & 0\\
04622 & RIG-I-like receptor signaling pathway & 38 & 0\\
05160 & Hepatitis C & 28 & 0\\
04940 & Type I diabetes mellitus & 35 & 0\\
\bottomrule
\end{tabular}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Create logical vector of whether in cluster 6}
\NormalTok{assignments =}\StringTok{ }\NormalTok{labels }\OperatorTok{==}\StringTok{ }\NormalTok{cluster}

@@

\begin{tabular}{llrrrlr}
\toprule
GO ID & Description & Annotated & Significant & Expected & P-value & Adj. p-value\\
\midrule
GO:0019395 & fatty acid oxidation & 55 & 54 & 44.78 & 0.00016 & 0.164\\
GO:1905114 & cell surface receptor signaling pathway ... & 272 & 243 & 221.46 & 0.00021 & 0.179\\
GO:0003002 & regionalization & 174 & 158 & 141.67 & 0.00049 & 0.358\\
GO:0009755 & hormone-mediated signaling pathway & 82 & 77 & 66.76 & 0.00105 & 0.671\\
GO:0048562 & embryonic organ morphogenesis & 154 & 139 & 125.39 & 0.00175 & 0.994\\
\addlinespace
GO:0006805 & xenobiotic metabolic process & 50 & 48 & 40.71 & 0.00257 & 1.000\\
GO:0090596 & sensory organ morphogenesis & 138 & 124 & 112.36 & 0.00332 & 1.000\\
GO:0035108 & limb morphogenesis & 74 & 69 & 60.25 & 0.00337 & 1.000\\
GO:0030166 & proteoglycan biosynthetic process & 27 & 27 & 21.98 & 0.00385 & 1.000\\
GO:0003206 & cardiac chamber morphogenesis & 72 & 67 & 58.62 & 0.00448 & 1.000\\
\addlinespace
GO:0055114 & oxidation-reduction process & 536 & 472 & 436.41 & 0.00489 & 1.000\\
GO:0021954 & central nervous system neuron developmen... & 36 & 35 & 29.31 & 0.00557 & 1.000\\
GO:0006261 & DNA-dependent DNA replication & 70 & 64 & 56.99 & 0.00595 & 1.000\\
GO:0006633 & fatty acid biosynthetic process & 78 & 70 & 63.51 & 0.00686 & 1.000\\
GO:0006790 & sulfur compound metabolic process & 154 & 136 & 125.39 & 0.00710 & 1.000\\
\bottomrule
\end{tabular}

@@

##  [3] ggfortify_0.4.10            timecoursedata_0.99.15     
